### 2018-06-01 CS761-RandomizedAlgorithms/assignments/a1.tex
q4

\newpage
\section{}
\end{document}

>>>>

\newpage
\section{}
\paragraph{}
Let $i > m$. Let $H(i)$ be the event that you hire candidate $i$. Let $B(i)$ be the event that candidate $i$ is the best candidate. Then $E_i = H(i) \cap B(i)$. So using Bayes formula we see that
$$Pr(E_i) = Pr(H(i) \mid B(i)) \cdot Pr(B(i)).$$
Now since the candidates are distributed uniformly at random $Pr(B(i)) = \frac{1}{n}$.
\begin{lemma}
Let $F(i)$ be the event that the best candidate among the first $i$ candidates appears in the first $m$ candidates. Then
$$Pr(H(i) \mid B(i)) = Pr(F(i-1)).$$
\end{lemma}
\begin{proof}
\paragraph{}
Given that $i$ is the best candidate, we hire $i$ if and only if we reject the first $i-1$ candidates. If we reject the first $i-1$ candidates then the best candidate among the first $i-1$ appears in the first $m-1$ candidates. Otherwise, it would not be rejected when it appears, as it is better than all previously seen candidates. On the other hand if the best candidates among the first $i-1$ appears in the first $m$ candidates, then we will reject the first $i-1$ candidates as none of them are better than a candidate which appears in the first $m$. Thus $H(i)$ happens given $B(i)$ if and only if $F(i-1)$ happens, i.e. the claim holds.
\end{proof}
\paragraph{}
Using this lemma we have
$$Pr(H(i) \mid B(i)) = Pr(F(i-1)) = \frac{m}{i-1}$$
since there are $m$ chances among $i-1$ that the best candidate among the first $i-1$ appears in the first $m$.
Therefore $Pr(E_i) = \frac{m}{n(i-1)}$.
\paragraph{}
Now the event $E$ is the disjoint union of events $E_1, \dots, E_n$. So we observe 
$$Pr(E) = Pr(\bigcup_{i} E_i) = \sum_{i} Pr(E_i) = \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1}$$
with the last equality following from our previous observations and the fact that $Pr(E_i) = 0$ for $i \leq m$.
\paragraph{}
We want to show that $Pr(E)$ can get arbitrarily close to $\frac{1}{e}$ for an appropriate choice of $m$. First we see that
$$Pr(E) =  \frac{m}{n}\sum_{i=m+1}^n \frac{1}{i-1} = \frac{m}{n}(\sum_{j=1}^{n-1}\frac{1}{j} - \sum_{j=1}^{m-1}\frac{1}{j}) =\frac{m}{n}(H_{n-1} - H_{m-1}).$$
Using that $\ln(n) \leq H_n \leq \ln(n+1)$ we have
$$Pr(E) \geq \frac{m}{n}(\ln(n-1) - \ln(m)) = \frac{m}{n}\ln(\frac{n-1}{m}).$$
If we choose $m = \frac{n-1}{e}$ then
$$Pr(E) \geq \frac{1}{e}\cdot\frac{n-1}{n}\ln(e) = \frac{1}{e}\cdot\frac{n-1}{n}$$
and taking the limit as $n$ tends to $\infty$ yields
$$\lim_{n\rightarrow \infty} Pr(E) \geq \frac{1}{e}.$$
\newpage
\section{}
\paragraph{}
Let $X$ be a random variable counting the number of cycles in a random permutation. Let $X_i$ be a random variable counting the number of cycles of length $i$ in a permutation. Then $X = \sum_i X_i$. So by linearity of expectation
$$E[X] = \sum_i E[X_i].$$
Now let $X_i^j$ be an indicator variable for $j \in {n\choose i}$ forms a length $i$ cycle in the random permutation. Then 
$$Pr(X^j_i) = \frac{(i-1)!(n-i)!}{n!}$$
since there are $n!$ permutations, and $(i-1)!(n-i)!$ permutations where $j$ forms a length $i$ cycle. To see this, consider a permutation where $j$ forms a length $i$ cycle. There are $(i-1)!$ ways to arrange the elements of $j$ in a length $i$ cycle, and $(n-i)!$ ways to permute the other elements. Therefore
$$E[X_i] = \sum_{j \in {n\choose i}}E[X_i^j] = {n\choose i} \frac{(i-1)!(n-i)!}{n!} = \frac{1}{i}.$$
Then
$$E[X] = \sum_{i=1}^n E[X_i] = \sum_{i=1}^n \frac{1}{i} = H_n.$$
\newpage
\section{}
\paragraph{}
We want to apply a Chernoff bound. So we will need to measure our run time in terms of some independent coin flips. If we can bound, for each element $x$ of our array, the depth of recursion before it is assigned a singleton array, then we can bound the runtime. Let $x$ be an element of our input array. Define $\ell_{x,i}$ to be the length of the array $x$ is assigned to at recursion depth $i$ of our quicksort algorithm. Note that our algorithm terminates at recursion depth $i$ if $\ell_{x,i}=1$ for all $x$.
\paragraph{}
Let $X_{x,i}$ be an indicator random variable indicating $1$ if $\ell_{x,i+1} \leq \frac{3}{4}\ell_{x,i}$, and taking value $0$ otherwise. Intuitively, $X_{x,i}$ indicates that we make a good split of the array containing $x$ at depth $i$ when quicksort is called on that array segment. Let $S_{x,i}$ be the array containing $x$ at recursion depth $i$. If our algorithm selects any element the interval $[\frac{\ell_{x,i}}{4}, \frac{3\ell_{x,i}}{4}]$ from the sorted version $S_{x,i}$ then $\ell_{x,i+1} \leq \frac{3}{4}\ell_{x,i}$. That means choosing one of at least $\frac{1}{2}$ the elements of $S$ as the pivot will yield to a good split, i.e.
$$Pr(X_{x,i}) \geq \frac{1}{2}.$$
So we can think of $X_{x,i}$ as a fair coin flip. Further each $X_{x,i+1}$ is independent of $X_{x,i}$ for all $i$.
\paragraph{}
Let $X_{x} = \sum_{i=1}^{k} X_{x,i}$ where $k=100\ln(n)$. Then if $X_{x} \geq d$ the length of the array containing $x$ at recursion depth $k$ is at most $\frac{3}{4}^d n$. If we want this to be a singleton then we need $$d = \frac{\ln(n)}{\ln{4/3}} \leq 4\ln(n) < 50\ln(n) = \frac{k}{2}$$
So now observing $E[X_{x}] =\frac{k}{2}$ we apply a Chernoff bound (for the lower tail, with $\mu = \frac{k}{2}$ and $\delta = \frac{1}{2}$)
$$Pr(X_{x} \leq \frac{1}{2}\frac{k}{2}) \leq e^{-\frac{k}{2}\frac{1}{8}} \leq e^{\frac{-50\ln(n)}{8}} \leq \frac{1}{n^6}$$
Applying the union bound we see that
$$Pr(\bigcup_x(X_{x} \leq \frac{1}{2}\frac{k}{2})) \leq \sum_{x} Pr(X_{x} \leq \frac{1}{2}\frac{k}{2})  \leq \frac{1}{n^5}.$$
Therefore with inverse polynomial probability, there is some element that is not in a singleton array after recursion depth of $100 \ln(n)$. Since each level of the recursion takes $n$ time, we see that with high probability the quicksort algorithm terminates in $100n\ln(n)$ time.
\end{document}
